Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-25 _tex/paper2.tex
Built site for gh-pages

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychometrika}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across different stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to have a higher trait level. For example, when assessing writing
quality, judges compare pairs of written texts (the stimuli) to
determine the relative writing quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of some traits or account for heterogeneous
stimuli, potentially leading to unreliable and inaccurate trait
estimates. Second, the study highlights the disconnect between CJ's
trait measurement and hypothesis testing processes. Although separating
these processes simplifies the analysis of CJ data, it may also
undermine the reliability of various statistical inferences derived from
these processes.

@@

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ assessment design features, such
as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Finally, Section~\ref{sec-discussion}
discusses the findings, explores avenues for future research, and detail
the challenges for future researchers.

@@

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions
\citep{Thurstone_1927b}. In Case I, Thurstone postulated that pairs of
stimuli would maintain a constant correlation across all comparisons. In
Case II, he allowed multiple judges to undertake comparisons instead of
confining evaluations to a single judge. In Case III, he posited that
there was no correlation between stimuli. In Case IV, he assumed that
the stimuli exhibited similar dispersions. Finally, in Case V, he
replaced this assumption with the condition that stimuli had equal
discriminal dispersions. Table~\ref{tbl-thurstone_cases} summarizes the
assumptions of the general form and the five cases. For a detailed
discussion of these cases and their progression, refer to
\citet{Thurstone_1927b} and \citet[pp.~248--253]{Bramley_2008}.

@@

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement and offering no guidance on how to use
its trait estimates for statistical inference or hypothesis testing.
Specifically, Thurstone cautioned that its use ``should not be made
without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as it
imposes the most extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to produce a ``rather coarse scaling'' of traits and ``allocate
the compared stimuli on this continuum''
\citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
statistical inference. Despite these limitations, it is surprising that
CJ research has predominantly relied on Case V to measure different
traits, which raises significant concerns about the reliability and
validity of such measurements in contexts where the case's assumptions
may not hold \citep{Kelly_et_al_2022, Andrich_1978}. Furthermore,
although the CJ tradition has attempted to address the gap of hypothesis
testing by using the point estimates of the traits--or their
transformations--a critical question remains: Does this approach provide
a valid foundation for statistical inference?

Thus, this section discusses these two prominent issues. Specifically,
Section~\ref{sec-theory-issue1} examines the heavy reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Conversely, Section~\ref{sec-theory-issue2} focuses on the apparent
disconnect between the approaches to trait measurement and hypothesis
testing in CJ.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep{Thurstone_1927a}.
Consequently, the outcome's ability to reflect \emph{true} differences
between stimuli largely depends on the validity of the model's
assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
equal dispersions. For instance, when the assumption accurately captures
the complexity of the data, the BTL model estimates a discriminal
difference distribution that accurately represents the \emph{true}
discriminal difference between the texts. This scenario is illustrated
in Figure~\ref{fig-dispersion}, when the model's discriminal difference
distribution aligns with the \emph{true} discriminal difference
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of this discriminal difference
then ensures reliable estimates for the texts' discriminal processes.

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail accommodate for heterogeneous stimuli, such as
handwritten texts or English compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
Indeed, the presence of the so-called \emph{misfit texts}--texts that
elicit more judgment discrepancies than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
signal this limitation, as such discrepancies can arise from larger
discriminal dispersions due to stimulus heterogeneity or because the
texts are genuine outliers--that is, texts with distinctive
characteristics that deviate markedly from the rest of the sample
\citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
it from adequately accounting for or addressing these anomalies, leaving
exclusion of the ``problematic'' texts as the primary remedy.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that \emph{misfit statistics} may represent texts with
different dispersions or outlying observations, the common CJ practice
of excluding stimuli based on these statistics may unintentionally
discard valuable information \citep{Miller_2023}, introducing bias into
the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which stimuli researchers exclude from the analysis.

@@

The correlation \(\rho\), illustrated in
Figure~\ref{fig-discriminal_difference}, measures how much the judges'
perception of a specific trait in one stimulus depends on their
perception of the same trait in another. Similar to the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
thought experiment, akin to the one presented in
Section~\ref{sec-theory-issue1a}, can illustrate this idea. Assuming
that the discriminal dispersions for both texts remain constant,
Figure~\ref{fig-correlation} shows that as the correlation between the
two texts increases, the distribution of their discriminal difference
becomes narrower. This narrowing, in turn, affects the probability that
the discriminal difference distinctly favors Text B over Text A--denoted
as \(P(B > A)\)--and thus directly influences the comparison outcome.
Furthermore, the figure shows that when two texts are independent or
uncorrelated, as assumed in the BTL model \((\rho=0)\), the distribution
of their discriminal difference is less narrow than in scenarios where
the texts are positively correlated. As a result, it becomes less likely
for the comparison to favor Text B over Text A, as indicated by the
larger shaded area.

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
assessment designs often adopt the assumption of no correlation between
stimuli based on an old theoretical justification. Specifically,
\citet{Thurstone_1927b} argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two scenarios: when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions or when additional hierarchical structures are
relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate complex, multidimensional traits or heterogeneous
stimuli, they often rely on a variety of intricate stimulus
characteristics to inform their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
Regardless of their relevance, these characteristics may not receive
equal weight or consistently oppose one another across comparisons. As a
result, they may exert a disproportionate influence on judges'
perceptions, generating biases that persist rather than cancel out. For
example, this could occur when a judge assessing the argumentative
quality of a text may place disproportionate emphasis on handwriting
clarity, thereby favoring neatly written texts despite their weaker
arguments. Moreover, because the discriminal process of stimuli becomes
an observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
scenario is limited, existing studies document the presence of judge
bias in CJ \citep[\citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2020a}]{Pollitt_et_al_2003}, reinforcing the
argument that the factors influencing pairwise comparisons do not always
cancel each other out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli--a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when the same individual
produces multiple texts, those texts often exhibit common features, such
as writing style or overall quality, that judges can easily recognize.
In this regard, although the CJ literature acknowledges the existence of
such hierarchical structures, the statistical approaches to account for
this additional source of dependence have been insufficient. For
instance, when CJ data incorporates multiple samples of stimuli from the
same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, similar to the ones discussed in
Section~\ref{sec-theory-issue2}.

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. Specifically,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023} and even biases
\citep{Wu_et_al_2022}. This issue is illustrated in
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the BTL scores follows the thick continuous line
\((\rho = 0)\), while the \emph{true} discriminal difference follows any
discontinuous line where \(\rho \neq 0\). Moreover, these inaccuracies
can result in spurious conclusions about the discriminal differences
\citep{McElreath_2020} and, by extension, the underlying discriminal
processes of the stimuli--potentially causing the model to incorrectly
classify stimuli or judges as \emph{misfits}.

Finally, similar to what we discussed in the previous section, removing
\emph{misfit} judges--that is, judges whose assessments deviate markedly
from the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and may appear as outliers under the BTL model
\citep{Wu_et_al_2022}--risk discarding valuable information and
introducing bias into trait estimates \citep{Miller_2023}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which judges researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Building on the previous section, it is clear that, researchers in CJ
studies typically use the BTL model to measure traits and position the
compared stimuli along a latent continuum \citep{Thurstone_1927b}. The
CJ literature also show that they frequently rely on point estimates of
these traits--typically the BTL scores or its transformations--to
conduct statistical inference or hypothesis testing. For example,
researchers have used these scores to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

In aggregate, the heavy reliance on Thurstone's Case V assumptions in
the statistical analysis of comparative data can compromise the
reliability of trait estimates. This overreliance may also undermine
their validity \citep{Perron_et_al_2015}, particularly when coupled with
the disconnect between the trait measurement and hypothesis testing
processes. However, the structural approach to causal inference can
address these issues by offering a systematic and integrated framework
that strengthens measurement reliability and validity while enhancing
the statistical accuracy of hypothesis tests.

@@

A motivating example that will appear throughout this document clarifies
these concepts. In this example, researchers aim to answer the question:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ assessment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''.

Following standard CJ practices, researchers typically use estimates
from the BTL model--or its transformations--to approximate this
estimand. However, as discussed in Section~\ref{sec-theory-issues},
Thurstone's Case V and the BTL model exhibit several statistical and
measurement limitations. These limitations hinder the model's ability to
identify various estimands relevant to CJ inquiries, including the one
described in the example.

>>>>

\usepackage{float}
\floatplacement{table}{H}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychometrika}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across different stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to have a higher trait level. For example, when assessing writing
quality, judges compare pairs of written texts (the stimuli) to
determine the relative writing quality each text exhibit (the trait)
\citep{Pollitt_2012b, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of some traits or account for heterogeneous
stimuli, potentially leading to unreliable and inaccurate trait
estimates. Second, the study highlights the disconnect between CJ's
trait measurement and hypothesis testing processes. Although separating
these processes simplifies the analysis of CJ data, it may also
undermine the reliability of various statistical inferences derived from
these processes
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ assessment design features, such
as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Finally, Section~\ref{sec-discussion}
discusses the findings, explores avenues for future research, and detail
the challenges for future researchers.

@@

@@

To address this issue and facilitate the practical implementation of the
theory, Thurstone developed five cases derived from this general form,
each progressively incorporating additional simplifying assumptions
\citep{Thurstone_1927b}. In Case I, Thurstone postulated that pairs of
stimuli would maintain a constant correlation across all comparisons. In
Case II, he allowed multiple judges to undertake comparisons instead of
confining evaluations to a single judge. In Case III, he posited that
there was no correlation between stimuli. In Case IV, he assumed that
the stimuli exhibited similar dispersions. Finally, in Case V, he
replaced this assumption with the condition that stimuli had equal
discriminal dispersions. Table~\ref{tbl-thurstone_cases} summarizes the
assumptions of the general form and the five cases. For a detailed
discussion of these cases and their progression, refer to
\citet{Thurstone_1927b} and \citet{Bramley_2008}.

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement and offering no guidance on how to use
its trait estimates for statistical inference or hypothesis testing.
Specifically, Thurstone cautioned that its use ``should not be made
without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, as it
imposes the most extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to produce a ``rather coarse scaling'' of traits and ``allocate
the compared stimuli on this continuum''
\citep[pp.~269]{Thurstone_1927b}, his theory did not support formal
statistical inference. Despite these limitations, CJ research has
predominantly relied on Case V to measure different traits, which raises
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
tradition has attempted to address the gap in hypothesis testing by
relying on the point estimates of traits--or their transformations--the
statistical literature cautions against using these estimates as the
sole basis for statistical inference, as such practices introduce bias
in the analysis and reduce the precision of hypothesis tests
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}. Next, both
issues are discussed in more depth.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep{Thurstone_1927a}.
Consequently, the outcome's ability to reflect \emph{true} differences
between stimuli largely depends on the validity of the model's
assumptions \citep{Kohler_et_al_2019}, in this case, the assumption of
equal dispersions. When the assumption accurately captures the
complexity of the data, the BTL model estimates a discriminal difference
distribution that accurately represents the \emph{true} discriminal
difference between the texts. This scenario is illustrated in
Figure~\ref{fig-dispersion}, when the model's discriminal difference
distribution aligns with the \emph{true} discriminal difference
distribution, represented by the thick continuous line corresponding to
\(\sigma_{B}-\sigma_{A}=0\). The accuracy of this discriminal difference
then ensures reliable estimates for the texts' discriminal processes.

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to accommodate for heterogeneous stimuli, such as
handwritten texts or English compositions
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.
The presence of the so-called \emph{misfit texts}--texts that elicit
more judgment discrepancies than others
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}--may
signal this limitation, as such discrepancies can arise from larger
discriminal dispersions due to stimulus heterogeneity or because the
texts are genuine outliers--that is, texts with distinctive
characteristics that deviate markedly from the rest of the sample
\citep{Grubbs_1969}. In either case, the BTL model's assumptions prevent
it from adequately accounting for or addressing these anomalies, leaving
exclusion of the ``problematic'' texts as the primary remedy
\citep{Pollitt_2012a, Pollitt_2012b}.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that \emph{misfit statistics} may represent texts with
different dispersions or outlying observations, the common CJ practice
of excluding stimuli based on these statistics may unintentionally
discard valuable information \citep{Miller_2023}, and introduce bias
into the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which stimuli researchers exclude from the analysis.

@@

The correlation between two stimuli \(\rho\) measures how much the
judges' perception of a specific trait in one stimulus depends on their
perception of the same trait in other stimulus. Similar to the
discriminal dispersions, this correlation shapes the distribution of the
discriminal difference, directly impacting the outcomes of pairwise
comparisons. Assuming that the discriminal dispersions for a couple of
texts remain constant, Figure~\ref{fig-correlation} shows that as the
correlation between the two texts increases, the distribution of their
discriminal difference becomes narrower. This narrowing, in turn,
affects the probability that the discriminal difference distinctly
favors Text B over Text A--denoted as \(P(B > A)\)--and thus directly
influences the comparison outcome. Furthermore, the figure shows that
when two texts are independent or uncorrelated, as assumed in the BTL
model \((\rho=0)\), the distribution of their discriminal difference is
less narrow than in scenarios where the texts are positively correlated.
As a result, it becomes less likely for the comparison to favor Text B
over Text A, as indicated by the larger shaded area.

Despite these notable differences in the distribution of the discriminal
difference under various correlational assumptions, in practice,
assessment designs often adopt the assumption of no correlation between
stimuli based on an old theoretical justification. Specifically,
\citet{Thurstone_1927b} argued that stimuli could be treated as
uncorrelated because judges' biases--arising from two opposing and
equally weighted effects occurring during the pairwise
comparisons--would cancel each other out. This idea was later formalized
by \citet{Andrich_1978}, who provided a mathematical demonstration of
this cancellation using the BTL model under the assumption of
discriminal processes with additive biases. However, evidence from the
CJ literature indicates that the assumption of zero correlation does not
hold in practice in at least two cases: when intricate aspects of
multidimensional, complex traits or heterogeneous stimuli influence
judges' perceptions or when additional hierarchical structures are
relevant to the stimuli.

Research on text quality assessments suggests that when judges evaluate
complex, multidimensional traits or heterogeneous stimuli, they often
rely on a variety of intricate stimulus characteristics to inform their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022}.
Regardless of their relevance, these characteristics may not receive
equal weight or consistently oppose one another across comparisons. As a
result, they may exert a disproportionate influence on judges'
perceptions, generating biases that persist rather than cancel out. For
example, this could occur when a judge assessing the argumentative
quality of a text may place disproportionate emphasis on handwriting
clarity, thereby favoring neatly written texts despite their weaker
arguments. Moreover, because the discriminal process of stimuli becomes
an observable outcome only through the judges' perceptions, these biases
could introduce dependencies between the stimuli
\citep{vanderLinden_et_al_2017_II}. While direct evidence for this exact
scenario is limited, existing studies document the presence of judge
bias in CJ
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
reinforcing the argument that the factors influencing pairwise
comparisons do not always cancel each other out.

In the second case, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli--a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. For instance, when the same individual
produces multiple texts, those texts often share several features such
as writing style or overall quality. Although the CJ literature
acknowledges the existence of such hierarchical structures
\citep[e.g.,][]{Boonen_et_al_2020}, the statistical approaches to
account for this additional source of dependence have been insufficient.
For instance, when CJ data incorporates multiple samples of stimuli from
the same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Thus, erroneously assuming zero correlation between stimuli can also
lead to significant statistical and measurement issues. In particular,
neglecting judges' biases or relevant hierarchical structures can create
dimensional mismatches in the model, leading to the over- or
underestimation of trait reliability
\citep{Ackerman_1989, Hoyle_et_al_2023} and even introduce statistical
biases \citep{Wu_et_al_2022}. These inaccuracies can result in spurious
conclusions about the discriminal differences \citep{McElreath_2020}
and, by extension, the underlying discriminal processes of the stimuli.
One such spurious conclusion could be the incorrect classification of
stimuli or judges as \emph{misfits}. Figure~\ref{fig-correlation}
illustrates how assuming zero correlation can undermine trait
reliability: the discriminal difference distribution of the BTL scores
follows the thick continuous line \((\rho = 0)\), while the \emph{true}
discriminal difference may correspond to any discontinuous line where
\(\rho \neq 0\).

Finally, removing \emph{misfit} judges--that is, judges whose
assessments deviate markedly from the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and may appear as outliers under the BTL model
\citep{Wu_et_al_2022}--risks discarding valuable information and
introducing bias into trait estimates \citep{Miller_2023}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which judges researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Researchers in CJ studies typically use the BTL model to measure traits
and position the compared stimuli along a latent continuum
\citep{Thurstone_1927b}. The CJ literature shows that research
frequently relies on point estimates of these traits--typically the BTL
scores or its transformations--to conduct statistical inference or
hypothesis testing. For example, researchers have used these scores to
identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

In aggregate, the heavy reliance on Thurstone's Case V assumptions in
the statistical analysis of comparative data can compromise the
reliability of trait estimates. This overreliance may also undermine
their validity \citep{Perron_et_al_2015}, particularly when coupled with
the disconnect between the trait measurement and hypothesis testing
processes. The structural approach to causal inference can address these
issues by offering a systematic and integrated framework to extend
Thurstone's general form. This approach can also strengthen measurement
reliability and validity while enhancing the statistical accuracy of
hypothesis tests.

@@

A motivating example that will appear throughout this study clarifies
these concepts. In this example, researchers aim to answer the question:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ assessment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''.

Following standard CJ practices, researchers would typically use
estimates from the BTL model--or its transformations--to approximate
this estimand. However, as discussed in Section~\ref{sec-theory-issues},
Thurstone's Case V and the BTL model exhibit several statistical and
measurement limitations. These limitations hinder the model's ability to
identify various estimands relevant to CJ inquiries, including the one
described in the example.

Fortunately, SCMs and DAGs support identification analysis through two
key advantages\footnote{In depth explanation of these topics is beyond
  the scope of this study, thus, readers seeking a more profound
  understanding can refer to introductory papers such as
  \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
  \citet{Cinelli_et_al_2020}, and introductory books like
  \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
  \citet{McElreath_2020} are useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2025}, are recommended.}. First, regardless of
complexity, they can represent various causal structures using only five
fundamental building blocks \citep{Neal_2020, McElreath_2024}. This
feature allows researchers to decompose complex structures into
manageable components, facilitating their analysis. Second, they depict
causal relationships in a non-parametric way. This flexibility enables
feasible identification strategies without requiring specification of
the types of variables, the functional forms relating them, or the
parameters of those functional forms \citep{Pearl_et_al_2016}.
